---
title: "Data Science Course Project"
output: html_notebook
---
Problem Statement:
To predict if a person have heart disease or not based
on attributes blood pressure,heart beat , 
exang = (exang: Exercise induced angina (1 = yes; 0 = no) )
fbs = (Fasting blood sugar (FBS) measures blood glucose after you have not eaten for at least 8 hours) (1 if blood sugar>120 mg) 
cp = chest pain (0<1<2<3)
ca = coronary calcium scan
trestbps = resting blood pressure in mmHg
chol = cholesterol (mg)
thalach = maximum heart rate
oldpeak = ST(points on ECG plot) depression induced by exercise relative to rest
restecg = rest Electro cardio graph
target = if there is heart disease it will show 1 otherwise it will show 0
```{r}
install.packages("tidyverse")
library(tidyverse)
data<-read.csv("C:/Users/shubh/Downloads/heart.csv")
View(data)
```


```{r}
head(data)
```
```{r}
tail(data)
```
```{r}
glimpse(data)
```
```{r}
ncol(data)
nrow(data)
```
```{r}
summary(data)
```
```{r}
data2<- data %>%
  mutate(sex=if_else(sex==1,"male","female"),
         fbs=if_else(fbs==1,">120","<=120"),
         exang=if_else(exang==1,"yes","no"),
         cp=if_else(cp==1,"ATYPICAL ANGINA",
                    if_else(cp==2,"NON-ANGINAL PAIN","ASYMPTOMATIC")),
         restecg=if_else(restecg==0,"NORMAL",
         if_else(restecg==1,"ABNORMALITY","PROBABLE OR DEFINITIVE")),
         slope=as.factor(slope),
         ca=as.factor(ca),
         thal=as.factor(thal),
         target=if_else(target==1 ,"YES","NO")
         ) %>%
  mutate_if(is.character,as.factor) %>%
  dplyr::select(target,sex,fbs,exang,cp,restecg,slope,ca,thal,everything())
  #install.packages("repr")  
  library(repr)
```

```{r}
# change the size of the plot
options(repr.plot.width =6 ,repr.plot.height = 3)
```

```{r}
#bar plot for target (heart disease)
ggplot(data2,aes(x=data2$target,fill=data2$target))+
  geom_bar() +
  xlab("Heart Disease") +
  ylab("Count") +
  ggtitle("Presence & Absence of Heart Disease") +
  scale_fill_discrete(name = 'Heart Disease', labels = c("Absent", "Present"))
```
```{r}
prop.table(table(data2$target))
```
```{r}
#count the frequency of the values of age

data2 %>%
  group_by(age) %>%
  count() %>%
  filter(n>10) %>%
  ggplot()+
  geom_col(aes(age,n),fill='green')+
  ggtitle("Age Analysis")+
  xlab("Age")+
  ylab("Age Count")
```
```{r}
#plot the distribution of the resting blood pressure
ggplot(data2,aes(x=trestbps))+
  geom_bar()

```
```{r}
#create a basic ggplot2 pie chart displaying the sex column of thne dataset
plotdata<- data2 %>%
  count(sex) %>%
  arrange(desc(sex)) %>%
  mutate(prop = round(n*100/sum(n),1),
         lab.ypos=cumsum(prop)- 0.5 *prop)
ggplot(plotdata,
       aes(x=" ",
           y=prop,
           fill=sex))+
  geom_bar(width=1, 
           stat = "identity",
           color="black")+
  coord_polar("y",
              start =0,
              direction = -1)+
  theme_void()
```
```{r}
#fasting blood sugar
plotdata<- data2 %>%
  count(fbs) %>%
  arrange(desc(fbs)) %>%
  mutate(prop=round(n*100/sum(n), 1),
         lab.ypos=cumsum(prop)- 0.5 * prop)
 ggplot(plotdata,
        aes(x=" ",
            y=prop,
            fill=fbs))+
   geom_bar(width=1,
            stat = "identity",
            color="black")+
   coord_polar("y",
               start = 0,
               direction = -1)+
   theme_void()
```
```{r}
#bivariate plots dependent on two variables for displaying the result

#cp: the chest pain experienced (value 1: typical angina, value 2: atypical angina , value 3: non anginal pain , value 4: asymptomatic)
ggplot(data2,
       aes(x=target,
           fill=cp ))+
  geom_bar(position = "stack")

```
```{r}
ggplot(data2,
       aes(x=sex,
           fill=cp))+
  geom_bar(position = "stack")
```
```{r}
#compare blood pressure across the chest pain by their respective genders

data2 %>%
  ggplot(aes(x=sex,y=trestbps))+
  geom_boxplot(fill='purple')+
  xlab("Sex")+
  ylab("BP")+
  facet_grid(~cp)
```
```{r}
#compare blood pressure across the chest pain irrespective of their genders
data2 %>%
  ggplot(aes(x=sex,y=trestbps))+
  geom_boxplot(fill="purple")+
  xlab("Sex")+
  ylab("BP")+
  facet_grid(~cp)
```
```{r}
#compare cholestrol across the chest pain by their respective genders
data2 %>%
  ggplot(aes(x=sex,y=chol))+
  geom_boxplot(fill="orange")+
  xlab("Sex")+
  ylab("Cholestrol")+
  facet_grid(~cp)
```
```{r}
#simple scatterplot
ggplot(data2, 
       aes(x=chol,
           y=age))+
  geom_point()
```

```{r}
#scatterplot with a linear progression direction 
#thalach : maximum heart rate achieved
ggplot(data2,
       aes(x=thalach,
           y=age))+
  geom_point(color="steelblue")+
  geom_smooth(method = "lm")
```
```{r}
#Scatterplot with a polynomial direction , comparing age with cholestrol
ggplot(data2,
       aes(x=age,
           y=chol))+
  geom_point(color="steelblue")+
  geom_smooth(method = "lm",
              formula = y~poly(x,2),
              color="indianred3")

```
```{r}
install.packages('caTools')
library(caTools)

#randomly split data
set.seed(123)
split=sample.split(data$target,SplitRatio = 0.75)
split

qualityTrain=subset(data,split==TRUE)
qualityTest=subset(data,split==FALSE)

```
```{r}
nrow(qualityTrain)
nrow(qualityTest)

```
```{r}
#building the model 
# the dependent variable used is target, for the independent variable is age , trestbps, chol, fbs , restecg, thalach,exang, oldpeak,slope
#e,ca,and thal.
#logistic regression model
data=glm(target ~
           age+trestbps+chol+fbs+
           restecg+thalach+exang+oldpeak+slope+ca+thal,
         data = qualityTrain,family = binomial)
summary(data)
```
```{r}
#a lot of variables are not significant.
#now we will removing variables based on Significance level
#using the backward method

data=glm(target~age+sex+cp+trestbps+chol+fbs+thalach+
           exang+oldpeak+slope+ca+thal,
         data=qualityTrain,family = binomial)
summary(data)
```
```{r}
#after removing as many as 4 variables in the following order :
#1.remove restecg
#2.remove fbs
#3.remove slope
#4.remove exang

data=glm(target~sex+cp+trestbps+chol+thalach+oldpeak+ca+thal,
         data=qualityTrain,family = binomial)
summary(data)
```
```{r}
#applying model after removing least significant variables
# a general rule in machine learning is that then more features you 
# have , the more likely model will suffer from overfitting

predictTrain=predict(data,type = "response")
predictTrain
```
```{r}
#plotting rocr curve 
#install.packages("ROCR")
library(ROCR)
rocrpred=prediction(predictTrain,qualityTrain$target)
rocrperf=performance(rocrpred,'tpr','fpr')

#plot(rocrperf,colorize=TRUE)
plot(rocrperf,colorize=TRUE,print.cutoffs.at=seq(0,1,by=0.1),
     text.adj=c(-0.2,1.7))
```
```{r}
#from rocr curve threshold of 0.7 seems to be okay so that true positives are maximised such that maximum number of patients with heart disease
# are not identified as healthy.
# then we can see the value of AUC , higher the AUC ,better the model is at distinguishing between patients with disease and no disease.
# area under then curve
auc= as.numeric(performance(rocrpred,'auc')@y.values)
auc
```
auc value is 0.91 that means our model is able to distinguish between patients with the disease and no disease with a probability of 0.91 so it is a good value 

```{r}
#accuracy using a threshold of 0.7

predictTest=predict(data,newdata = qualityTest,type = 'response')
table(qualityTest$target,predictTest >=0.7)
```
```{r}
#accuracy 
(113+90)/257

```
#logistic regression model with all the variables and logistic regression model after removing less significant attributes performed best with an accuracy of testing 78%
```{r}
# Load required packages
library(caret)

# Load heart disease data
data<-read.csv("C:/Users/shubh/Downloads/heart.csv")


# Fit logistic regression model
model <- train(target ~ ., data = data, method = "glm", family = "binomial")

# Function to validate numeric input
validate_numeric_input <- function(x) {
  while (is.na(as.numeric(x))) {
    x <- readline(prompt = "Invalid input. Please enter a numeric value: ")
  }
  return(as.numeric(x))
}

# Prompt user for input values
age <- validate_numeric_input(readline(prompt = "Enter your age: "))
sex <- validate_numeric_input(readline(prompt = "Enter your sex (0 = female, 1 = male): "))
cp <- validate_numeric_input(readline(prompt = "Enter your chest pain type (0-3): "))
trestbps <- validate_numeric_input(readline(prompt = "Enter your resting blood pressure: "))
chol <- validate_numeric_input(readline(prompt = "Enter your serum cholesterol : "))
fbs <- validate_numeric_input(readline(prompt = "Enter your fasting blood sugar (0 = <= 120 mg/dL, 1 = > 120 mg/dL): "))
restecg <- validate_numeric_input(readline(prompt = "Enter your resting electrocardiographic results (0-2): "))
thalach <- validate_numeric_input(readline(prompt = "Enter your maximum heart rate achieved : "))
exang <- validate_numeric_input(readline(prompt = "Enter whether you have exercise-induced angina (0 = no, 1 = yes): "))
oldpeak <- validate_numeric_input(readline(prompt = "Enter your ST depression induced by exercise relative to rest: "))
slope <- validate_numeric_input(readline(prompt = "Enter the slope of the peak exercise ST segment (0-2): "))
ca <- validate_numeric_input(readline(prompt = "Enter the number of major vessels colored by fluoroscopy (0-3): "))
thal <- validate_numeric_input(readline(prompt = "Enter your thalassemia (0-3): "))
# Create data frame with input values
new_data <- data.frame(age = age, sex = sex, cp = cp, trestbps = trestbps, chol = chol, fbs = fbs, restecg = restecg,
                       thalach = thalach, exang = exang, oldpeak = oldpeak, slope = slope, ca = ca, thal = thal)

# Use model to predict target
prediction_prob <- predict(model, newdata = new_data, type = "raw")

# Output prediction
# Use model to predict target probabilities
prediction_prob <- predict(model, newdata = new_data, type = "raw")

# Output prediction with probability
if (prediction_prob >= 0.5) {
  message("You are predicted to have heart disease with a probability of ", round(prediction_prob, 2))
} else {
  message("You are predicted to not have heart disease with a probability of ", round(1-prediction_prob, 2))
}
```



